\name{}: Supplement data

### 2021-06-29 paper/main.tex
Draft for cover letter response in main paper, added heatmap and interaction table

% With the rapidly growing field of machine learning approaches and
% their application to bioscientifical problems in the realm of
% bioinformatics, different kinds of data, such as long DNA sequences
% could be utilized for feature generation, while rapid advances were
% made. 
Identifying drug--target interactions (DTIs) is important in drug
repurposing and repositioning, and in the exploration of novel drug
candidates and their interaction partners. Computational methods are
widely applied to predict DTIs and many computational methods have
been developed.  These methods can be broadly classified into
``top-down'' and ``bottom-up'' approaches. Top-down approaches start
from observable characteristics resulting from a drug--target
interaction, such as side-effects or the diseases treated by a drug,
and infer likely molecular mechanisms (i.e., the interaction) using
these observations.
% and indications represented by knowledge graphs or ontologies, induced by a drug and infer targets based on the likely molecular mechanisms that result in these phenotypes
Bottom-up approaches start from molecular features such as molecular
structure or fingerprints associated with drug and protein, and
predict interactions from this information.

Both bottom-up and top-down approaches to drug--target interaction
prediction bear some advantages and limitations. Generally, bottom-up
methods face the challenge to predict whether a structure binds to a
protein given their molecular properties; whether two entities
interact depends not only on the molecular structure of the entities
(where binding sites and molecular forces need to be determined for
accurate prediction) but also properties namely where a protein is
expressed or in which celltypes and anatomical structures. Top-down
methods use information for DTI prediction such as side-effect
similarity \cite{Campillos2008} that is largely complementary to
knowledge gained from molecular properties. While methods that rely on
molecular information are directly predicting whether two molecular
could interact, top-down methods base on more indirect means and 
infer a DTI from observable effects resulting from the interaction.

@@

@@

Another view on these two approaches is as direct and indirect ways to
predict drug--target interactions. On one hand, molecular information can be used
to directly determine whether two molecules (such as a drug and
protein) have the ability to interact, whereas information about
phenotypic consequences of a drug (drug effects) or disruption of a
protein function can be used to indirectly suggest candidate
drug--target interactions. Molecular features will be specific to a
drug--target pair and we would not expect this information to
propagate through a protein--protein interaction network; the main
information about drug--target interactions that could be obtained
from interactions between proteins is information about binding sites
between proteins that may also be used by a drug molecule (i.e.,
information that protein $P_1$ binds to protein $P_2$ reveals
information about the molecular structures of both $P_1$ and
$P_2$). On the other hand, phenotypic consequences of changes in
protein function, or drug effects, are often a result of aberrant
pathway or network activity and involve more than one protein;
consequently, we expect these features to benefit more from including
information about protein--protein interactions. Our results (Table
\ref{tab:Results}) confirm this hypothesis and demonstrate that
molecular features do not benefit from including the interaction
network whereas the indirect, top-down features benefit from the
propagation over the interaction network.

@@

The most common scheme for drug--target interaction prediction is the
split over drug-target pairs \cite{Survey2018} where it may happen
that most drugs and targets that are including in the model's
validation and testing phase have also been included in the training
phase (as part of other drug--target pairs). This scheme is prone to a
number of biases. If the number of interactions for a drug or protein
are imbalanced, i.e., some drugs or proteins have many more
interactions than others, these will be seen more often during
training and they will likely also have more interactions in the
testing and validation sets; because some entities have more
interactions, i.e., they are more likely to interact, any model that
preferentially predicts these as interaction partners will improve its
predictive performance. While this accurately captures the
distribution, predicting based on biases in the number of interaction
partners is not desirable when applying the model to novel
entities. We have demonstrated that even a ``na\"ive'' classifier that
makes predictions only by exploiting the imbalanced number of
interaction partners can achieve performance close to state of the art
methods (when measuring Macro AUC). When training a machine learning
model on such imbalanced data, it will eventually overfit to this
imbalance.  Splitting by entity (protein or drug) can reduce the
impact of this spurious correlation but not reduce it entirely,
because similar entities will still exhibit similar interaction
patterns. In our experiments, we observed the impact of splitting
training and evaluation sets by protein as a decrease in overall
performance (Macro AUC), providing some evidence that models trained
using this splitting scheme are less sensitive to overfitting to this
type of bias.

@@

>>>>

% With the rapidly growing field of machine learning approaches and
% their application to bioscientifical problems in the realm of
% bioinformatics, different kinds of data, such as long DNA sequences
% could be utilized for feature generation, while rapid advances were
% made. 
Identifying drug--target interactions (DTIs) is one of the initial steps in drug discovery and screening for disease treatment and thus helps to mitigate the the costs and risks of drug development.  Further, it forms a generalization of both drug repurposing and repositioning, finding novel functions for already approved and extensively analyzed drugs with well explored properties. Inferring and deriving from those existing properties and interactions can help to analyze and identify novel drugs for known targets, leading to desirable therapeutic effects. Moreover, while \textit{in vitro} DTI prediction requires the availability of all substances, computational, \textit{in silico} DTI predictors interpret and anticipate millions of interactions within short time.

Computational methods are
widely applied to predict DTIs and many computational methods have
been developed.  These methods can be broadly classified into
``top-down'' and ``bottom-up'' approaches. Top-down approaches start
from observable characteristics resulting from a drug--target
interaction, such as side-effects or the diseases treated by a drug,
and infer likely molecular mechanisms (i.e., the interaction) using
these observations.
% and indications represented by knowledge graphs or ontologies, induced by a drug and infer targets based on the likely molecular mechanisms that result in these phenotypes
Bottom-up approaches start from molecular features such as molecular
structure or fingerprints associated with drug and protein, and
predict interactions from this information.

Both bottom-up and top-down approaches to drug--target interaction
prediction bear some advantages and limitations. Generally, bottom-up
methods face the challenge to predict whether a chemical structure binds to a
protein given their molecular properties; whether two entities
interact depends not only on the molecular structure of the entities
(where binding sites and molecular forces need to be determined for
accurate prediction) but also properties namely in which celltypes and anatomical structures a protein is expressed. Top-down
methods use information for DTI prediction such as side-effect
similarity \cite{Campillos2008} that is largely complementary to
knowledge gained from molecular properties. While methods that rely on
molecular information are directly predicting whether two molecules
could interact, top-down methods base on more indirect means and 
infer a DTI from observable effects resulting from the interaction.

@@

We further investigate the influence of the more specific interaction types \textbf{inhibition} and \textit{activation} within STITCH database on the performance of \name{}. The results are summarized in Supplementary table 3, where we find no difference in comparison to the complete dataset.

@@

Another view on these two approaches is as direct and indirect ways to
predict drug--target interactions. On one hand, molecular information can be used
to directly determine whether two molecules (such as a drug and
protein) have the ability to interact, whereas information about
phenotypic consequences of a drug (drug effects) or disruption of a
protein function can be used to indirectly suggest candidate
drug--target interactions. Molecular features will be specific to a
drug--target pair and we would not expect this information to
propagate through a protein--protein interaction network; the main
information about drug--target interactions that could be obtained
from interactions between proteins is information about binding sites
between proteins that may also be used by a drug molecule (i.e.,
information that protein $P_1$ binds to protein $P_2$ reveals
information about the molecular structures of both $P_1$ and
$P_2$). On the other hand, phenotypic consequences of changes in
protein function, or drug effects, are often a result of aberrant
pathway or network activity and involve more than one protein;
consequently, we expect these features to benefit more from including
information about protein--protein interactions. Moreover, as the protein--protein 
interaction relation is not transitive (protein $P_1$ interacting with $P_2$, 
and $P_2$ relating with $P_3$ does not imply $P_1$ interacting
or influencing $P_3$), we can only infer from the direct neighborhood of each protein within the PPI graph. 
Our results (Table
\ref{tab:Results}) confirm the first hypothesis and demonstrate that
molecular features do not benefit from including the interaction
network whereas the indirect, top-down features benefit from the
propagation over the interaction network. While the exploitation of ontology based features over PPI graphs was already examined in \citet{Trebacz2020} for gene expression analysis, our work is the first to investigate and apply this technique to DTI prediction. Further, \name{} is the first work to combine both ``bottom-up'' and ``top-down'' features.

@@

The most common scheme for drug--target interaction prediction is the
split over drug-target pairs \cite{Survey2018} where it may happen
that most drugs and targets that are including in the model's
validation and testing phase have also been included in the training
phase (as part of other drug--target pairs). This scheme is prone to a
number of biases. If the number of interactions for a drug or protein
are imbalanced, i.e., some drugs or proteins have many more
interactions than others, these will be seen more often during
training and they will likely also have more interactions in the
testing and validation sets; because some entities have more
interactions, i.e., they are more likely to interact, any model that
preferentially predicts these as interaction partners will improve its
predictive performance. While this accurately captures the
distribution, predicting based on biases in the number of interaction
partners is not desirable when applying the model to novel
entities. We have demonstrated that even the newly proposed ``na\"ive'' classifier that
makes predictions only by exploiting the imbalanced number of
interaction partners can achieve performance close to state of the art
methods (when measuring Macro AUC). When training a machine learning
model on such imbalanced data, it will eventually overfit to this
imbalance.  Splitting by entity (protein or drug) can reduce the
impact of this spurious correlation but not reduce it entirely,
because similar entities will still exhibit similar interaction
patterns. In our experiments, we observed the impact of splitting
training and evaluation sets by protein as a decrease in overall
performance (Macro AUC), providing some evidence that models trained
using this splitting scheme are less sensitive to overfitting to this
type of bias.

@@

\subsection{Pharmacological novelty}

As the target-based predictive power of \name{} improves significantly over other methods, we utilized our model to predict novel drug classes for protein families. We therefore collected the 2nd level ATC (Anatomical Therapeutic Chemical) groups for each drug and all InterPro \cite{Interpro2020} top-level families for each protein. Utilizing the STITCH interactions, we followed a protein split within each InterPro family by predicting over all available drugs. We eventually normalized the number of novel interactions per class by the amount of proteins within the respective family. A heatmap showing the results of this analysis can be found in Supplementary figure 5. \dots \todo{Roberts class analysis}.

### 2021-06-29 paper/supplement/supplement.tex
.
Draft for cover letter response in main paper, added heatmap and interaction table

\begin{table}[ht]
	\centering
	\begin{tabular}{|p{2.0cm}|p{0.8cm}|p{0.8cm}|p{0.8cm}|p{0.8cm}|p{0.8cm}|p{0.8cm}|}
		\hline
		&\multicolumn{6}{c|}{(a) STITCH results}\\
		\hline
		\name{} results&\multicolumn{6}{c|}{PPI graph}\\
		&\multicolumn{3}{c|}{without}&\multicolumn{3}{c|}{with}\\
		&Macro AUC&Micro $AUC'_p$&Micro $AUC_p$&Macro AUC&Micro $AUC'_p$&Micro $AUC_p$\\
		\hline
		MolPred&$0.69$&$0.66$&$0.65$&$0.69$&$0.68$&$0.67$\\
		\hline
		OntoPred&$0.88$&$0.89$&$0.87$&$0.92$&$0.94$&$0.93$\\
		\hline
		\name{} & $0.89$& $0.91$ & $0.90$&$\mathbf{0.93}$&$\mathbf{0.94}$&$\mathbf{0.94}$\\
		\hline
	\end{tabular}
	\caption{Extended results table for MolPred, OntoPred and
          \name{} over STITCH with added $MicroAUC'_p$ for comparison
          of both metrics.}
\end{table}

@@

	\begin{tabular}{|p{2.0cm}|p{1.1cm}|p{1.2cm}|p{0.8cm}|p{0.8cm}|p{1.2cm}|p{1.2cm}|}
		\hline
		Approach&\multicolumn{6}{c|}{Graph convolution method}\\
		&without&without/ empty RGC&GCN-Conv&GEN-Conv&RGC + GCNConv&RGC + GENConv\\
		\hline
		MolPred & $0.69$ & $0.67$& $0.67$& $0.68$& $0.65$& $0.69$\\
		\hline
		OntoPred &$0.88$ &$0.87$ &$0.88$&$0.90$&$0.86$&$0.92$\\
		\hline
		\name& $0.89$&$0.89$& $0.89$& $0.91$& $0.86$& $\mathbf{0.93}$\\
		\hline
	\end{tabular}
	\caption{\label{tab:Results}Results for various neural graph
          convolutional methods enhancing MolPred, OntoPred and
          \name{} over the STITCH dataset. We evaluated all methods
          over different stacking heights without and with residual
          graph convolution blocks (RGC, ResGraphConv, see
          Supplementary figure \ref{fig:ResGraphConvBlocks}). ``Empty
          RGC'' denotes a RGC block with a plain neural layer
          substituting the GraphConv layer.}
\end{table}

>>>>

\begin{table}[ht]
	\centering
	\begin{tabular}{|p{2.0cm}|p{0.8cm}|p{0.8cm}|p{0.8cm}|p{0.8cm}|p{0.8cm}|p{0.8cm}|}
		\hline
		&\multicolumn{6}{c|}{(a) STITCH results}\\
		\hline
		\name{} results&\multicolumn{6}{c|}{PPI graph}\\
		&\multicolumn{3}{c|}{without}&\multicolumn{3}{c|}{with}\\
		&Macro AUC&Micro $AUC'_p$&Micro $AUC_p$&Macro AUC&Micro $AUC'_p$&Micro $AUC_p$\\
		\hline
		MolPred&$0.69$&$0.66$&$0.65$&$0.69$&$0.68$&$0.67$\\
		\hline
		OntoPred&$0.88$&$0.89$&$0.87$&$0.92$&$0.94$&$0.93$\\
		\hline
		\name{} & $0.89$& $0.91$ & $0.90$&$0.93$&$0.94$&$0.94$\\
		\hline
	\end{tabular}
	\caption{Extended results table for MolPred, OntoPred and
          \name{} over STITCH with added $MicroAUC'_p$ (without
          imputation) for comparison of both metrics.}
\end{table}

@@

	\begin{tabular}{|p{2.0cm}|p{1.1cm}|p{1.2cm}|p{0.8cm}|p{0.8cm}|p{1.2cm}|p{1.2cm}|}
		\hline
		Approach&\multicolumn{6}{c|}{Graph convolution method (MacroAUC)}\\
		&without&without/ empty RGC&GCN-Conv&GEN-Conv&RGC + GCNConv&RGC + GENConv\\
		\hline
		MolPred & $0.69$ & $0.67$& $0.67$& $0.68$& $0.65$& $0.69$\\
		\hline
		OntoPred &$0.88$ &$0.87$ &$0.88$&$0.90$&$0.86$&$0.92$\\
		\hline
		\name& $0.89$&$0.89$& $0.89$& $0.91$& $0.86$& $\mathbf{0.93}$\\
		\hline
	\end{tabular}
	\caption{\label{tab:Results}Results for various neural graph
          convolutional methods enhancing MolPred, OntoPred and
          \name{} over the STITCH dataset. We evaluated all methods
          over different stacking heights without and with residual
          graph convolution blocks (RGC, ResGraphConv, see
          Supplementary figure \ref{fig:ResGraphConvBlocks}). ``Empty
          RGC'' denotes a RGC block with a plain neural layer
          substituting the GraphConv layer.}
\end{table}

\clearpage
\begin{figure}[ht]%figure1
	\centerline{\includegraphics[width=1.1\textwidth]{../figures/ATC_Interpro_heatmap.png}}
	\caption{Heatmap showing the results of the ATC drug class and InterPro family prediction analysis}
	\label{fig:ATCInterproHeatmap}
\end{figure}

\clearpage
\begin{table}[ht]
	\centering
	\begin{tabular}{|p{2.0cm}|p{1cm}|p{1cm}|p{1cm}|p{1cm}|}
		\hline
		&\multicolumn{4}{c|}{Interaction type}\\
		&\multicolumn{2}{c|}{Activation}&\multicolumn{2}{c|}{Inhibition}\\
		\#Drugs & \multicolumn{2}{l|}{723} & \multicolumn{2}{l|}{954}\\
		\#Proteins&\multicolumn{2}{l|}{4562}&\multicolumn{2}{l|}{6222}\\
		\#Links&\multicolumn{2}{l|}{14174}&\multicolumn{2}{l|}{16364}\\
		\hline
		Metric&Macro AUC&Micro AUC$_p$&Macro AUC&Micro AUC$_p$\\
		\name&$0.95$&$0.95$&$0.93$&$0.94$\\
		\hline

	\end{tabular}
	\caption{Extended results table for MolPred, OntoPred and
		\name{} over STITCH with added $MicroAUC'_p$ (without
		imputation) for comparison of both metrics.}
\end{table}
